Viscoelastic Kelvin-Helmholtz instability

### 2013-10-22 paper/story/story.tex
adding initial version of paper

%	paper.tex	Toby Searle
%	Last modified: Tue 22 Oct 13:14:21 2013
%	using jfm template

\NeedsTeXFormat{LaTeX2e}

\documentclass{jfm}

\usepackage{graphicx}
\usepackage{natbib}
%%%%%%%%%%%Added to template:
\usepackage{caption,subcaption, mathtools} % add my own packages
\graphicspath{ {./images/} }
\newcommand\Wi{\mbox{\textit{Wi}}}
\newcommand{\dt}[1]{\frac{d #1}{d t}} %time deriv
\newcommand{\dy}[1]{\frac{\partial #1}{\partial y}}
%%%%%%%%%%%%%%%

% See if the author has AMS Euler fonts installed: If they have, attempt
% to use the 'upmath' package to provide upright math.
\ifCUPmtlplainloaded \else
  \checkfont{eurm10}
  \iffontfound
    \IfFileExists{upmath.sty}
      {\typeout{^^JFound AMS Euler Roman fonts on the system,
                   using the 'upmath' package.^^J}%
       \usepackage{upmath}}
      {\typeout{^^JFound AMS Euler Roman fonts on the system, but you
                   dont seem to have the}%
       \typeout{'upmath' package installed. JFM.cls can take advantage
                 of these fonts,^^Jif you use 'upmath' package.^^J}%
       \providecommand\upi{\pi}%
      }
  \else
    \providecommand\upi{\pi}%
  \fi
\fi

% See if the author has AMS symbol fonts installed: If they have, attempt
% to use the 'amssymb' package to provide the AMS symbol characters.

\ifCUPmtlplainloaded \else
  \checkfont{msam10}
  \iffontfound
    \IfFileExists{amssymb.sty}
      {\typeout{^^JFound AMS Symbol fonts on the system, using the
                'amssymb' package.^^J}%
       \usepackage{amssymb}%
       \let\le=\leqslant  \let\leq=\leqslant
       \let\ge=\geqslant  \let\geq=\geqslant
      }{}
  \fi
\fi

% See if the author has the AMS 'amsbsy' package installed: If they have,
% use it to provide better bold math support (with \boldsymbol).

\ifCUPmtlplainloaded \else
  \IfFileExists{amsbsy.sty}
    {\typeout{^^JFound the 'amsbsy' package on the system, using it.^^J}%
     \usepackage{amsbsy}}
    {\providecommand\boldsymbol[1]{\mbox{\boldmath $##1$}}}
\fi

%%% Example macros (some are not used in this sample file) %%%

% For units of measure
\newcommand\dynpercm{\nobreak\mbox{$\;$dyn\,cm$^{-1}$}}
\newcommand\cmpermin{\nobreak\mbox{$\;$cm\,min$^{-1}$}}

% Various bold symbols
\providecommand\bnabla{\boldsymbol{\nabla}}
\providecommand\bcdot{\boldsymbol{\cdot}}
\newcommand\biS{\boldsymbol{S}}
\newcommand\etb{\boldsymbol{\eta}}

% For multiletter symbols
\newcommand\Real{\mbox{Re}} % cf plain TeX's \Re and Reynolds number
\newcommand\Imag{\mbox{Im}} % cf plain TeX's \Im
\newcommand\Rey{\mbox{\textit{Re}}}  % Reynolds number
\newcommand\Pran{\mbox{\textit{Pr}}} % Prandtl number, cf TeX's \Pr product
\newcommand\Pen{\mbox{\textit{Pe}}}  % Peclet number
\newcommand\Ai{\mbox{Ai}}            % Airy function
\newcommand\Bi{\mbox{Bi}}            % Airy function

% For sans serif characters:
% The following macros are setup in JFM.cls for sans-serif fonts in text
% and math.  If you use these macros in your article, the required fonts
% will be substitued when you article is typeset by the typesetter.
%
% \textsfi, \mathsfi   : sans-serif slanted
% \textsfb, \mathsfb   : sans-serif bold
% \textsfbi, \mathsfbi : sans-serif bold slanted (doesnt exist in CM fonts)
%
% For san-serif roman use \textsf and \mathsf as normal.
%
\newcommand\ssC{\mathsf{C}}    % for sans serif C
\newcommand\sfsP{\mathsfi{P}}  % for sans serif sloping P
\newcommand\slsQ{\mathsfbi{Q}} % for sans serif bold-sloping Q

% Hat position
\newcommand\hatp{\skew3\hat{p}}      % p with hat
\newcommand\hatR{\skew3\hat{R}}      % R with hat
\newcommand\hatRR{\skew3\hat{\hatR}} % R with 2 hats
\newcommand\doubletildesigma{\skew2\tilde{\skew2\tilde{\Sigma}}}
%       italic Sigma with double tilde

% array strut to make delimiters come out right size both ends
\newsavebox{\astrutbox}
\sbox{\astrutbox}{\rule[-5pt]{0pt}{20pt}}
\newcommand{\astrut}{\usebox{\astrutbox}}

\newcommand\GaPQ{\ensuremath{G_a(P,Q)}}
\newcommand\GsPQ{\ensuremath{G_s(P,Q)}}
\newcommand\p{\ensuremath{\partial}}
\newcommand\tti{\ensuremath{\rightarrow\infty}}
\newcommand\kgd{\ensuremath{k\gamma d}}
\newcommand\shalf{\ensuremath{{\scriptstyle\frac{1}{2}}}}
\newcommand\sh{\ensuremath{^{\shalf}}}
\newcommand\smh{\ensuremath{^{-\shalf}}}
\newcommand\squart{\ensuremath{{\textstyle\frac{1}{4}}}}
\newcommand\thalf{\ensuremath{{\textstyle\frac{1}{2}}}}
\newcommand\Gat{\ensuremath{\widetilde{G_a}}}
\newcommand\ttz{\ensuremath{\rightarrow 0}}
\newcommand\ndq{\ensuremath{\frac{\mbox{$\partial$}}{\mbox{$\partial$} n_q}}}
\newcommand\sumjm{\ensuremath{\sum_{j=1}^{M}}}
\newcommand\pvi{\ensuremath{\int_0^{\infty}%
  \mskip \ifCUPmtlplainloaded -30mu\else -33mu\fi -\quad}}

\newcommand\etal{\mbox{\textit{et al.}}}
\newcommand\etc{etc.\ }
\newcommand\eg{e.g.\ }

\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

\title[Viscoelastic Kelvin-Helmholtz instabilty]{Viscoelastic Kelvin-Helmholtz instability}

\author[T. W. Searle and A. N. Morozov]%
{T. W. Searle$^1$ and A. N. Morozov$^1$%
  \thanks{Email address for correspondence: jfm@damtp.cam.ac.uk},\ns
}

% NOTE: A full address must be provided: department, university/institution, town/city, zipcode/postcode, country.
\affiliation{$^1$SUPA, School of Physics and Astronomy, University of Edinburgh, Mayfield Road,
Edinburgh, EH9 3JZ, UK\\[\affilskip]
}

\pubyear{2013}
\volume{650}
\pagerange{119--126}
% Do not enter received and revised dates. These will be entered by the editorial office.
\date{?; revised ?; accepted ?. - To be entered by editorial office}
%\setcounter{page}{1}
\begin{document}

@@

\begin{abstract}
  Abstract goes here. Abstract goes here. Viscoelastic Kelvin-Helmholtz instability. 
\end{abstract}

\begin{keywords}
Authors should not enter keywords on the manuscript, as these must be chosen by the author during the online submission process and will then be added during the typesetting process (see http://journals.cambridge.org/data/\linebreak[3]relatedlink/jfm-\linebreak[3]keywords.pdf for the full list)
\end{keywords}

\section{Introduction}

@@

In order to simulate this flow, we used an Oldroyd-B fluid with very low Reynolds number $\Rey < 0.01$ and a small ratio between the solvent and total viscosities $\beta = \frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid, where the polymer contribution to the dynamics is higher than the solvent contribution. The Oldroyd-B Navier stokes and constitutive equations are then:
\begin{align}
    Re \left[ \dt{\mathbf{v}} + \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T} + \nabla{\mathbf{v}}
\end{align}
With $Wi$ denoting the Weissenberg number and $\tau$ the stress in the polymeric fluid. The FENE-CR viscoelastic fluid has the advantage of being similarly easy to simulate, but also including a finite extensibility for the polymers. This leads to a stress which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells via a non-linear spring force:

@@

\begin{figure}
    \centering
    \includegraphics[width=0.3\textwidth]{KH_diagram}
    \caption{diagram of the system}
    \label{fig:diagram}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{KH_dispersion}
    \caption{Example of dispersion relation for Weissenberg numbers 200, 100 and $10^{-5}$, with $\beta = 0.1$ and $\Rey = 10000$. Polymer elasticity reduces the height of the instability.}
    \label{fig:KH_growth_rate}
\end{figure}

We performed a Chebyshev decomposition of the flow in the wall-normal ($y$) direction for the base profile, and decomposed the disturbances to this flow in both Chebyshev and Fourier modes.

\begin{equation}
    g(y) = \sum\limits_{m=1}^{m=M} \widetilde{g}_{m} T_{m}(y) e^{ikx + \lambda t}
\end{equation}

for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $M$ is the number of Chebyshev modes, $T_{m}$ is the mth Chebyshev polynomial of the 1st kind, $k$ is the streamwise wavenumber of the disturbance and $\lambda$ is the growth rate of the mode.

\section{A purely elastic instability}

@@

The instability remains unchanged under variation of $\Delta$ at low $\Delta$, proving that it is a free shear instability. Although the instability appears to require some inertia in the fluid, it is still present for  $\Rey = 0.01$  - far lower than the usual threshold of $\Rey = 1000$ for a Newtonian Kelvin-Helmholtz instability.

We see an instability grow from around $k_x \sim 0.1$ and move to lower $k_x$ as the Weissenberg number increases. The height of the instability also increases with increasing Weissenberg number.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_purely_elastic}
	\caption{}
	\label{fig:inf_low_Re}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_dispersions_low_Re}
	\caption{}
	\label{fig:dispersions_low_Re}
    \end{subfigure}
    \caption{a) Free shear version of the instability. Plot of the maximum growth rate against Wiessenberg number at $\Rey = 0.01$. b) Dispersion relations at various Weissenberg numbers to accompany the trend in figure a).}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_vary_Re}
	\caption{}
	\label{fig:inf_low_Re}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_dispersions_low_Wi}
	\caption{}
	\label{fig:dispersions_low_Wi}
    \end{subfigure}
    \caption{a) Free shear version of the instability. Plot of the maximum growth rate against Reynold's number with $\Wi=50$. b) Dispersion relations at various Reynold's numbers to accompany the trend in figure a).}
\end{figure}

\subsection{ Shear instability with walls}

If we examine the eigenvectors of the instability, we see that there are very large polymer stresses in the $\pm \Delta$ region. This corresponds to the large shear rate brought about as the two flows pass each other. Examining the flow field, we see that there are large vortices arranged with opposite rotational directions above and below the shear region which extend almost to the walls.

\begin{figure}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{psi_low_delta}
    \caption{}
    \label{fig:psi_low_delta}
    \end{subfigure}
    \vspace{1 mm}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{txx_low_delta}
    \caption{}
    \label{fig:stress_low_delta}
    \end{subfigure}
    \caption{a) Streamfunction of the largest eigenfunction of the instability when $\Rey = 0.1$ and $\Wi =100$. b) xx stress for the same instability.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{KH_vary_delta}
    \caption{Dispersion relations for various values of $\Delta$ at $\Rey = 0.1$, $\Wi = 100$. The sensitivity to the walls seems only to increase the strength of the instability. Even when the system is insensitive to the walls, at $\Delta= 0.0125$, there is still an instability present.}
    \label{fig:walls_dependence}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{psi_high_delta}
    \caption{Streamfunction of the eigenfunction of the instability at $\Delta = 0.1$. The magnitude is scaled by the xx stress in the centre of the channel.}
    \label{fig:psi_high_delta}
\end{figure}

All of this goes to say why it is difficult to see the instability at $\Rey = 0$. As the Reynold's number is reduced the walls become more important to the instability. In order to remove the dependence on the walls and recover a free shear instability it is necessary to move them further away. This problem gets worse at lower $\Rey$. By introducing just a very small contribution from inertia, we can remove the dependence on the walls. This is why all of our results are taken for $\Rey \leq 0.1$ on the length scale of the instability.

The reason we have have the simulation with the walls at all is because of the implications for the plane Couette problem. The Kelvin-Helmholtz instability in the Newtonian problem takes place on a length scale of about 10\% of the width of the channel, or $\Delta = 0.1$ (a guess). The viscoelastic Kelvin-Helmholtz instability is sensitive to the walls at this $\Delta$ and a $\Rey \lesssim 1$ (see figure \ref{fig:vary_Re_delta_conv}). This means that on the scale of the channel, $\Rey \gtrsim 10$ for the walls not to affect the instability. So, the important results for the purely elastic instability from the point of view of the plane Couette problem are probably those for which the walls are relevant.

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{vary_Re_delta_conv}
    \caption{The blue line is the data from the stretched code and the green line is the $\Delta = 0.1$ data. By $\Rey = 1$ it is clear that the Reynold's number is sufficient such that the walled system behaves as though it were insensitive to the walls.}
    \label{fig:vary_Re_delta_conv}
\end{figure}

At $\Delta = 0.1$ we see the same instability as that seen in the completely free case.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
    \includegraphics[width=\textwidth]{KH_purely_elastic}
    \caption{}
    \label{fig:KH_purely_elastic}
\end{subfigure}
~
\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_low_Wi_vary_Re}
    \caption{}
    \label{fig:KH_reduce_Re}
\end{subfigure}

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_low_Re_vary_Wi}
    \caption{ }
    \label{fig:KH_purely_elastic}
\end{subfigure}
~
\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
\caption{a) The change in the height of the dispersion relation as $\Wi$ is increased at $\Rey = 0$. This instability must be entirely elastic in nature. b) The change in the height of the dispersion relation as the Reynolds number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still present for large polymer concentrations, consistent with purely elastic turbulence. c) The change in the height of the dispersion relation as $\Wi$ is increased at $\Rey = 0$. This instability must be entirely elastic in nature. d) The value of the Weissenberg number at which the velocity profile becomes unstable against $\beta$. The onset of the purely elastic instability moves to higher $Wi$ as $\beta$ increases.}
\end{figure}

\section{Confirming the model}

\subsection{Consistency of results with the FENE model}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{FENE_low_Re_vary_Wi}
    \caption{Results are consistent with figure \ref{fig:KH_purely_elastic}}
    \label{fig:FENE_low_Re}
\end{figure}

\subsection{Varying width of the instability}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{high_Re_vary_delta}
    \caption{When results are rescaled for the various delta, The instability is found to be insensitive to the width of the instability relative to the wall. At delta = 0.1, $\Rey = 10000$, $\Wi = 0.001$. 300 Gauss-Labatto points were used for each calculation.}
    \label{fig:delta_dispersion}
\end{figure}

\subsection{Drag reduction}

We see drag reduction using the Oldroyd-B model. It is not clear whether or not the maximum drag reduction asymptote is present.

\begin{figure}
%    \showthe\columnwidth %use this to determine size of figures.
    \centering
    \includegraphics[width=0.7\textwidth]{KH_high_Re_vary_Wi}
    \caption{The height of the dispersion relation is reduced for the Kelvin-Helmholtz instability as the Weissenberg number is increased. The grey dashed line is the height of the dispersion relation for the Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased stability of the base flow and so confirms the presence of drag reduction in this system.}
    \label{fig:KH_drag_reduction}
\end{figure}

\section{Conclusions}

@@

\bibliographystyle{jfm}
% Note the spaces between the initials

%\bibliography{jfm-instructions}

>>>>

%
%   The Story of a Kelvin-Helmholtz like viscoelastic instability
%   Last modified: Mon 21 Oct 11:00:34 2013

%   PREAMBLE:
\documentclass{article}
\usepackage{amsmath, amssymb}
%   MY MATH MACROS 
\newcommand{\dt}[1]{\frac{d #1}{d t}} %time deriv
\newcommand{\dy}[1]{\frac{\partial #1}{\partial y}}

\begin{document}
\title{Viscoelastic Kelvin-Helmholtz instability}

@@

@@

In order to simulate this flow, we used an Oldroyd-B fluid with very low Reynolds number $Re < 0.01$ and a small ratio between the solvent and total viscosities $\beta = \frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid, where the polymer contribution to the dynamics is higher than the solvent contribution. The Oldroyd-B Navier stokes and constitutive equations are then:
\begin{align}
    Re \left[ \dt{\mathbf{v}} + \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T} + \nabla{\mathbf{v}}
\end{align}
With $Wi$ denoting the Weissenberg number and $\tau$ the stress in the polymeric fluid. The FENE-CR viscoelastic fluid has the advantage of being similarly easy to simulate, but also including a finite extensibility for the polymers. This leads to a stress which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells via a non-linear spring force:

@@

\section{Drag reduction}

We see a reduction in the height of the instability with increasing Weissenberg number at high Reynolds number and high $\beta$. This is consistent with the addition of a small amount of polymeric fluid reducing the instability of a Newtonian Kelvin-Helmholtz flow. Presumably, this would bring about drag reduction on the walls and delay the onset of the turbulence to larger Reynold's number. This is further confirming evidence for the model when the Reynold's number is high.

\section{The Purely Elastic Instability}

@@

The instability remains unchanged under variation of $\Delta$ at low $\Delta$, proving that it is a free shear instability. Although the instability appears to require some inertia in the fluid, it is still present for  $Re = 0.01$  - far lower than the usual threshold of $Re = 1000$ for a Newtonian Kelvin-Helmholtz instability.

We found that the instability sets in at approximately $Wi = 15$ for $\beta = 0.1$, and that this onset Weissenberg number stays the same up to about $\beta = 0.6$ after which it quickly rises. This suggests that there is a concentration of polymers required in the fluid in order to enable the viscoelastic version of the instability.

@@

If we examine the eigenvectors of the instability, we see that there are very large polymer stresses in the $\pm \Delta$ region. This corresponds to the large shear rate brought about as the two flows pass each other. Examining the flow field, we see that there are large vortices arranged with opposite rotational directions above and below the shear region.

\section{Varying the distance of the channel walls}

@@

All of this goes to say why it is difficult to see the instability at $Re = 0$. As the Reynold's number is reduced the walls become more important to the instability. In order to remove the dependence on the walls and recover a free shear instability it is necessary to move them further away. This problem gets worse at lower $Re$. By introducing just a very small contribution from inertia, we can remove the dependence on the walls. This is why all of our results are taken for $\Re \leq 0.1$ on the length scale of the instability.

The reason we have have the simulation with the walls at all is because of the implications for the plane Couette problem. The Kelvin-Helmholtz instability in the Newtonian problem takes place on a length scale of about 10\% of the width of the channel, or $\Delta = 0.1$. The viscoelastic Kelvin-Helmholtz instability is sensitive to the walls at this $\Delta$ and a $Re \lesssim 10$. This means that on the scale of the channel $Re \gtrsim 100$ for the walls not to affect the instability. So, the important results for the purely elastic instability from the point of view of the plane Couette problem are probably those for which the walls are relevant.

\section{Consistency with the FENE-CR model}

@@

\section{Conclusion}

@@
